---
title: "Widget System Risk Analysis Report"
author: ""
date: '2024-08-15'
date-modified: '2025-08-18'
categories: []
order: 100
format:
  html:
    code-fold: true
    code-tools: true
    code-link: true
output:
  html_notebook:
    theme:
      version: 5
      preset: bootstrap
    pandoc_args: --shift-heading-level-by=1
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

Example quantrr risk analysis of the "Widget Management System" (WMS).

```{r setup, message = FALSE, warning = FALSE}
library(quantrr)
library(readxl)
library(janitor)
library(validate)
library(dplyr)
library(formattable)
library(purrr)
library(ggplot2)
library(plotly)

# TODO: workaround for https://github.com/r-lib/lintr/issues/2790, update when
#   https://github.com/data-cleaning/validate/pull/197 is released (1.1.6+)
is.nzchar <- nzchar # nolint: object_name_linter.

# set the relative file path and name to import
report_file <- "data/widgetsys.xlsx"
```

# Environment Statement

The Widget Management System (WMS) is over 30 years old and its architecture has not changed significantly since the original implementation. Over the years, the widget system has become an integral part of our services in managing widgets for our clients. In reviewing the system, three major risks were identified: First, the age of the technology prevents updating components of the system that no longer meet contemporary cybersecurity standards, which increases the likelihood of a breach. Second, the system is less reliable and experiences frequent outages, typically about 2 major outages per year, which results in lost revenue, contractual penalties, and overtime pay to recover from the incident. Third, limitations of the widget system have started to affect sales - we have recently lost a customer due to the functional obsolescence of the widget system, and expect to both lose more existing and prospective customers in the future due to increased competition in the widget management market.

Experts were asked to consider two risk treatment scenarios:

- None: The current system as it is today (baseline risk)
- Replace: Complete replacement of the WMS with a modern, customer-centric solution

# Import

Import and validate data from Excel.

The data was collected from 3 Technology SMEs, 3 Business SMEs, and one SME with experience in both. Experts were calibrated, informed by historical and industry data, and only gave estimates for areas in which they were confident in answering.

```{r import}
risks <- read_xlsx(report_file, sheet = "Risks") |>
  clean_names()

validate_risks <- local({
  validate_rules <- validator(
    risk_char = is.character(risk),
    risk_not_na = !is.na(risk),
    risk_not_blank = is.nzchar(risk, keep_na = TRUE),
    desc_char = is.character(description),
    desc_not_na = !is.na(description),
    desc_not_blank = is.nzchar(risk, keep_na = TRUE)
  )
  confront(risks, validate_rules)
})

check_validation(validate_risks, sheet = "Risks")

treatments <- read_xlsx(report_file, sheet = "Treatments") |>
  clean_names()

validate_treatments <- local({
  validate_rules <- validator(
    treatment_char = is.character(treatment),
    treatment_not_na = !is.na(treatment),
    treatment_not_blank = is.nzchar(treatment, keep_na = TRUE),
    desc_char = is.character(description),
    desc_not_na = !is.na(description),
    desc_not_blank = is.nzchar(treatment, keep_na = TRUE)
  )
  confront(treatments, validate_rules)
})

check_validation(validate_treatments, sheet = "Treatments")

estimates <- read_xlsx(report_file, sheet = "Estimates") |>
  clean_names() |>
  rename(
    lambda = frequency_per_yer, p05 = low_5_percent, p95 = high_95_percent, p50 = most_likely
  )

validate_estimates <- local({
  validate_rules <- validator(
    risk_not_na = !is.na(risk),
    risk_match = risk %in% risks$risk,
    expert_char = is.character(expert),
    expert_not_na = !is.na(expert),
    expert_not_blank = is.nzchar(expert, keep_na = TRUE),
    lambda_num = is.numeric(lambda),
    lambda_pos = lambda > 0,
    p05_num = is.numeric(p05),
    p05_pos = p05 > 0,
    p95_num = is.numeric(p95),
    p95_pos = p95 > 0,
    p50_num = is.numeric(p50),
    p50_pos = p50 > 0
  )
  confront(estimates, validate_rules)
})

check_validation(validate_estimates, sheet = "Estimates")
```

# Risks

Risk descriptions:

```{r risks}
formattable(risks, align = "l")
```

Treatment descriptions:

```{r treatments}
formattable(treatments, align = "l")
```

# Forecast

Forecast risk using Monte Carlo simulation. The average events, losses, 'typical' losses (geometric
mean), and percentage of years with no losses for each risk and treatment are summarized below:

```{r forecast}
consensus <- estimates |>
  group_by(treatment, risk) |>
  summarize(across(lambda:p50, ~ mean(.x, na.rm = TRUE)), .groups = "drop")

consensus_params <- consensus |>
  mutate(as_tibble(lnorm_param(.data$p05, .data$p95, .data$p50)))

forecast <- consensus_params |>
  select(c("risk", "lambda", "meanlog", "sdlog", "treatment")) |>
  pmap(calc_risk) |>
  list_rbind()

forecast |>
  group_by(treatment, risk) |>
  mutate(no_losses = events == 0) |>
  summarize(
    avg_events = mean(events), avg_losses = mean(losses), typ_losses = gmean(losses),
    no_losses = mean(no_losses), .groups = "drop"
  ) |>
  mutate(
    across(avg_losses:typ_losses, ~ currency(.x, digits = 0L)),
    no_losses = percent(no_losses)
  ) |>
  formattable(align = "l")
```

# Losses

Baseline losses (with no risk treatment) by risk:

```{r risk_hist}
forecast |>
  filter(tolower(treatment) == "none") |>
  # remove zero losses to plot using log10 scale
  filter(losses > 0) |>
  ggplot(aes(losses)) +
  facet_grid(vars(risk)) +
  geom_histogram(color = "black", fill = "white", bins = 100) +
  scale_x_log10(labels = scales::label_currency(scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = NULL, title = "Losses by Risk") +
  theme_minimal()
```

Compare risk treatments:

```{r treatment_hist}
forecast |>
  group_by(treatment, year) |>
  summarize(total_losses = sum(losses), .groups = "drop_last") |>
  # remove zero losses to plot using log10 scale
  filter(total_losses > 0) |>
  ggplot(aes(total_losses)) +
  facet_grid(vars(treatment)) +
  geom_histogram(color = "black", fill = "white", bins = 100) +
  scale_x_log10(labels = scales::label_currency(scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = NULL, title = "Total Losses by Treatment") +
  theme_minimal()
```

# Loss Exceedance Curves

Plot loss exceedance curves for all risks and combined risk, with risk treatments.

## By Risk

Plot loss exceedance curves for each risk:

```{r risk_le}
risk_le <- forecast |>
  group_by(risk, treatment) |>
  mutate(probability = 1 - percent_rank(losses)) |>
  filter(losses > 0) |>
  ungroup() |>
  mutate(losses = currency(losses, digits = 0), probability = percent(probability)) |>
  ggplot(aes(losses, probability, color = treatment)) +
  facet_grid(vars(risk)) +
  geom_line() +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
  scale_x_log10(labels = scales::label_currency(scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = NULL, color = "Treatment", title = "Loss Exceedance by Risk") +
  scale_color_viridis_d() +
  theme_minimal()

risk_le
```

Interactive plot:

```{r risk_le_plotly}
ggplotly(risk_le + guides(color = "none"))
```

## Combined Risk

Plot loss exceedance curve for combined risk:

```{r combined_le}
combined_le <- forecast |>
  group_by(treatment, year) |>
  summarize(total_losses = sum(losses), .groups = "drop_last") |>
  mutate(probability = 1 - percent_rank(total_losses)) |>
  filter(total_losses > 0) |>
  ungroup() |>
  mutate(total_losses = currency(total_losses, digits = 0), probability = percent(probability)) |>
  ggplot(aes(total_losses, probability, color = treatment)) +
  geom_line() +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
  scale_x_log10(labels = scales::label_currency(scale_cut = scales::cut_short_scale())) +
  labs(x = NULL, y = NULL, color = "Treatment", title = "Combined Loss Exceedance") +
  scale_color_viridis_d() +
  theme_minimal()

combined_le
```

Interactive plot:

```{r combined_le_plotly}
ggplotly(combined_le + guides(color = "none"))
```

# Appendix

Additional details on the risk quantification analysis.

## Validation

Data validation results for Risks tab:

```{r validate_risks}
plot(validate_risks)
```

Data validation results for Treatments tab:

```{r validate_treatments}
plot(validate_treatments)
```

Data validation results for Estimates tab:

```{r validate_estimates}
plot(validate_estimates)
```

## Estimates

All risk estimates:

```{r estimates}
estimates |>
  mutate(across(p05:p50, ~ currency(.x, digits = 0L))) |>
  formattable(align = "l")
```

## Consensus Estimate

Using a simple average of all experts that provided an estimate (not blank/NA), this gives us a
consensus estimate for the three risks of:

```{r consensus}
consensus |>
  mutate(across(p05:p50, ~ currency(.x, digits = 0L))) |>
  formattable(align = "l")
```

The consensus estimates for p05 and p95 result in the following parameters for log-normal loss
magnitude. The p50 estimate is used to calculate the percentage difference from the actual median
(`mdiff`), a measure of estimate accuracy:

```{r consensus_params}
consensus_params |>
  mutate(across(p05:p50, ~ currency(.x, digits = 0L)), mdiff = percent(mdiff)) |>
  formattable(align = "l")
```

## Forecast Summary

A `summary()` of the forecast results.

```{r summary}
summary(forecast)
```
